\documentclass[sigconf,nonacm]{acmart}

\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\title{LiveSafeRust: LSP-Guided Incremental Security Analysis for Rust}

\author{Anonymous Author(s)}
\affiliation{\institution{Anonymous Institution}\country{}}
\email{}

\begin{document}

\begin{abstract}
Rust developers receive fast feedback from the compiler and rust-analyzer, but
cross-procedure security analyses such as value-flow taint and typestate checks
typically remain outside the live editing loop. This paper proposes
LiveSafeRust, an orchestration layer above SAF that caches SAF-derived facts,
invalidates them with the editor's semantic changes, and republishes findings as
LSP diagnostics. We patched SAF to support \texttt{rustc}-emitted LLVM IR for
\texttt{std::process::Command} and path-sensitive file-opening cases. On seven
minimized Rust cases across command injection and path traversal, including a
minimized RUSTSEC-2024-0350 / CVE-2024-35186-style case, SAF reports 4 TP / 0 FP
/ 0 FN / 3 TN; LiveSafeRust agrees with the direct SAF verdict and serves a
semantic no-op edit from cached SAF JSON in 0.57 ms without re-invoking SAF. The
key idea is not a new taint engine, but a query boundary and diagnostic refresh
rule that amortize expensive SAF calls across edits. We state a diagnostic
refresh theorem showing when caller-only invalidation by summary inequality is
sufficient, describe conservative handling for unresolved Rust calls, and report
a preliminary sidecar prototype. A synthetic fan-out workload with 5131
functions validates the theorem's propagation cutoff: a one-line sanitizer
removal recomputes 122 summaries, refreshes five diagnostics, and agrees with
full reanalysis. On duct, cargo check takes 5.80 s cold and 0.05 s warm; the
same LiveSafeRust orchestration runs over 94 extracted functions in 0.6 ms after
an edit.
\end{abstract}

\maketitle

\section{Introduction}

Modern Rust tooling has made type errors, borrow checker feedback, and many
lint-like diagnostics feel nearly live. Security-oriented cross-procedure
analyses have not reached the same interaction model. Batch tools can reason
about value flow and resource protocols, but their workflow is poorly matched
to an editor loop where developers expect actionable feedback after each small
edit.

This paper asks whether SAF-derived findings can be kept responsive inside the
Rust LSP stack. Our hypothesis is that rust-analyzer's existing Salsa
recomputation graph is the right invalidation boundary: when a file changes,
rust-analyzer already invalidates parse, macro, HIR, and type queries. Security
summaries and cached SAF facts can be expressed as downstream queries over those
facts.

We make three contributions:
\begin{itemize}
  \item a LiveSafeRust orchestration architecture that keeps SAF-derived facts
  responsive inside an LSP loop;
  \item a diagnostic refresh theorem for summary-based IDE analyses;
  \item a preliminary sidecar prototype that patches and invokes SAF on
  Rust-generated LLVM IR for \texttt{std::process::Command}, then demonstrates
  cache/invalidation on SAF JSON.
\end{itemize}

\section{Gap And Motivation}

Rust has strong live tooling, but the live feedback boundary mostly ends at
syntax, type checking, borrow checking, and lint-like rules. Analyses that need
cross-procedure value-flow or typestate summaries are typically run as batch
tools or CI jobs. This creates a latency and workflow gap: diagnostics arrive
after the developer has moved away from the edit that introduced the issue.

We anchor this gap with initial measurements on the duct crate. Cargo check
takes 5.80 s cold and 0.05 s warm, while cargo clippy takes 2.24 s cold and
0.10 s warm. A minimal rust-analyzer LSP probe still needs work: it observed
empty diagnostics quickly, but did not produce a nonempty diagnostic for a
semantic-error edit in our minimal client.
This suggests LiveSafeRust's security stream is additive to, rather than
redundant with, rust-analyzer's own diagnostics budget.

\section{Method}

\subsection{Salsa Query Boundary}

LiveSafeRust treats security analysis artifacts as tracked queries:
\begin{itemize}
  \item \texttt{saf\_body\_facts(db, f, cfg)} lowers rust-analyzer HIR, source
  maps, and inference facts to sources, sinks, sanitizers, parameters, and call
  sites.
  \item \texttt{saf\_callees(db, f, cfg)} resolves call targets where possible.
  \item \texttt{taint\_summary(db, f, cfg)} computes the summary visible to
  callers.
  \item \texttt{local\_taint\_findings(db, f, cfg)} computes diagnostics whose
  physical sink is in the function body.
  \item \texttt{taint\_diagnostics(db, file, cfg)} aggregates findings into LSP
  diagnostics.
\end{itemize}

The summary contains caller-visible facts such as return dependencies and
sink-parameter dependencies. Diagnostics are not stored in summaries; they are
local products that depend on body facts and direct callee summaries.

\subsection{Theorem 1: Diagnostic Refresh By Summary Inequality}

\textbf{Theorem.} Under a finite monotone summary lattice and finding locality,
caller-only invalidation triggered by callee summary inequality is sufficient
for diagnostic completeness.

\textbf{Assumptions.} Each function summary is an element of a finite monotone
lattice. Local findings for function $f$ depend only on $f$'s current body facts
and summaries of functions directly called by $f$.

\textbf{Refresh rule.} After an edit, recompute summaries and local findings for
edited functions. Enqueue callers only when a recomputed callee summary changes.
Stop propagation at a caller whose recomputed summary equals its previous
summary.

\textbf{Proof sketch.} A function's ancestors observe the function only through
its summary. If the summary is unchanged, all transfer effects visible to
ancestors are unchanged; therefore no ancestor diagnostic can change through
that path. Local findings are refreshed because their query depends directly on
local body facts and direct callee summaries. Thus diagnostics whose truth value
could change are recomputed before publication, while ancestors behind an equal
summary need not be revisited.

\subsection{Rust-Specific Conservatism}

Rust introduces trait-object calls, function pointers, closures, macro-expanded
calls, and FFI boundaries. When rust-analyzer cannot resolve a call target with
enough confidence, LiveSafeRust uses a conservative top summary: the return may
depend on unknown taint and every argument may reach an unknown sink. Such
diagnostics should be marked as low confidence until refined by crate-specific
models or lower-level SAF facts.

\subsection{SAF Interface Boundary}

LiveSafeRust is an orchestration layer above SAF. SAF remains responsible for
computing taint facts and findings from IR; LiveSafeRust is responsible for
cache keys, edit invalidation, summary equality, and LSP publication. The
cache key is a semantic LLVM-IR hash that strips debug/source noise and
alpha-renames local SSA value names within each function while retaining global
function identifiers and call structure. The prototype invokes SAF through the Python SDK, stores the result as
\texttt{livesaferust.saf-backed/0.1} JSON, and feeds that JSON into the same
reverse-call-graph worklist used by the synthetic and syn-backed workloads.
Unsupported Rust IR shapes are reported directly rather than hidden behind the
source-sidecar model.

\section{SAF Patch}

The initial LLVM-22 SAF frontend parsed host \texttt{rustc 1.95} IR syntax but
crashed while lowering Rust aggregate constants used by
\texttt{std::process::Command}. The crash was triggered by array constants whose
LLVM textual form was \texttt{undef}/\texttt{poison}; the frontend called string
accessors on those values. We patched the LLVM mapping layer to conservatively
lower such arrays to SAF \texttt{Undef}. The value-flow builder then needed two
Rust ABI repairs: direct def-use edges for aggregate \texttt{memcpy} moves and a
conservative \texttt{sret} model that connects a call result and non-output
arguments to the first out-pointer operand. A regression fixture,
\texttt{tests/fixtures/llvm/e2e/rust\_std\_command.ll}, is covered by
\texttt{crates/saf-frontends/tests/rust\_std\_command\_frontend.rs} and
\texttt{crates/saf-analysis/tests/rust\_std\_command\_e2e.rs}. The full
LLVM-22 test suite passes after the patch (2193 Rust tests and 94 Python SDK
tests).

\section{Prototype}

Our current prototype is a sidecar rather than a rust-analyzer integration. It
generates Rust-like workloads, extracts Rust facts with \texttt{syn}, invokes
SAF on LLVM IR, caches the resulting facts, and emits LSP-shaped diagnostics.
The prototype is not evidence of full Rust semantic coverage; it is evidence
that the proposed SAF orchestration and invalidation policy are implementable
and measurable.

The main workload contains 5000 unrelated helper functions plus five independent
24-hop command-construction branches. A one-line edit removes the sanitizer in
the shared \texttt{normalize\_command} function.

\begin{table}[h]
\caption{Preliminary sidecar result. Times are JavaScript microbenchmark times,
not rust-analyzer or SAF-core times.}
\label{tab:prelim}
\begin{tabular}{lrrrrrr}
\toprule
Scenario & Funcs & Changed & Recomp. & Sum $\Delta$ & Stable & Findings \\
\midrule
Clean cold & 5131 & 5131 & 5131 & 5007 & 130 & 0 \\
Unrelated incr. & 5131 & 1 & 1 & 0 & 1 & 0 \\
Vulnerable incr. & 5131 & 1 & 122 & 121 & 1 & 5 \\
Vulnerable full & 5131 & 5131 & 5131 & 5128 & 9 & 5 \\
\bottomrule
\end{tabular}
\end{table}

The vulnerable incremental row illustrates Theorem 1. The edit causes 121
summary changes along the fan-out wrapper graph. The only stable recomputed
summary is the root caller, where propagation stops. Its five local diagnostics
are still refreshed and agree with full reanalysis.
This synthetic experiment is deliberately not the SAF-integration proof. It
stresses Theorem~1's summary-equality cutoff over a deep call graph. The SAF
experiment below validates a different part of the claim: that results produced
by the real SAF engine can be cached, invalidated, and published through the
LiveSafeRust diagnostic path. The current SAF SDK returns final findings rather
than per-function intermediate summaries, so the two experiments are
complementary. A unified implementation would expose SAF's per-function
summaries to the same orchestration layer.

\section{Evaluation Plan}

We measured duct, a 2134-line command-oriented Rust crate. The syn sidecar
extracts 94 functions/methods and one \texttt{Command::new} sink from
\texttt{src/lib.rs} in 0.04 s warm time.

\begin{table}[h]
\caption{Initial real-crate baseline on duct.}
\label{tab:duct-baseline}
\begin{tabular}{llr}
\toprule
System & Scenario & Time \\
\midrule
cargo check & cold & 5.80 s \\
cargo check & warm & 0.05 s \\
cargo check & warm after Clippy & 0.28 s \\
cargo clippy & cold & 2.24 s \\
cargo clippy & warm & 0.10 s \\
syn sidecar & warm fact extraction & 0.04 s \\
\bottomrule
\end{tabular}
\end{table}

The same duct facts now run through LiveSafeRust's summary engine. We apply a
compiling source edit, changing \texttt{Vec::new()} to
\texttt{Vec::<OsString>::new()} in \texttt{cmd}. The edit changes one function
hash but not the command sink facts.

\begin{table}[h]
\caption{LiveSafeRust summary analysis over duct syn facts. duct exposes zero
source hints and one \texttt{Command::new} sink, so zero findings is the
expected ground truth. Times exclude syn parsing and are JavaScript prototype
times.}
\label{tab:duct-livesaferust}
\begin{tabular}{lrrrrr}
\toprule
Scenario & Funcs & Changed & Recomp. & Findings & Time \\
\midrule
Cold & 94 & 94 & 94 & 0 & 1.8 ms \\
Edit incremental & 94 & 1 & 1 & 0 & 0.6 ms \\
Edit full & 94 & 94 & 94 & 0 & 1.0 ms \\
\bottomrule
\end{tabular}
\end{table}

The rust-analyzer LSP latency probe needs a fuller client/configuration before
it can be used as a nonempty diagnostics baseline. The next implementation step
is replacing the syn fact adapter with rust-analyzer HIR facts.

To show the diagnostic surface, we also compiled and analyzed a 7-line minimized
Rust case where \texttt{env::args()} flows directly into
\texttt{Command::new}. The syn sidecar extracts one function, one source hint,
and one sink hint; LiveSafeRust emits one LSP-shaped
\texttt{tainted-command} diagnostic in 0.2 ms. This case is not an efficacy
benchmark; it is an artifact for showing the trace payload that would appear in
an editor.

We then invoked the patched SAF directly on \texttt{std::process::Command}
Rust IR. The primary case sends \texttt{env::args()} into
\texttt{Command::new}; a second vulnerable case sends the same input into
\texttt{Command::arg}. We also added a second vulnerability family:
\texttt{env::args()} flowing into \texttt{std::fs::File::open}/
\texttt{File::create}, including a minimized RUSTSEC-2024-0350 style checkout
case where an untrusted repository path is joined with a worktree path before
file creation~\cite{rustsec-gixfs}.

\begin{table}[h]
\caption{SAF-backed Rust integration after the patch. Compile time is host
\texttt{rustc 1.95}; SAF time is measured inside the LLVM-22 Docker service.}
\label{tab:saf-backed}
\begin{tabular}{lrrrr}
\toprule
Case & Compile & SAF & Funcs & Findings \\
\midrule
\texttt{Command::new} vuln. & 0.38 s & 0.96 s & 166 & 1 \\
\texttt{Command::arg} vuln. & 0.04 s & 0.65 s & 171 & 1 \\
\texttt{File::open} traversal & 0.38 s & 0.79 s & 115 & 1 \\
RUSTSEC-2024-0350 min. & 0.38 s & 0.81 s & 135 & 1 \\
Clean print & 0.04 s & 0.58 s & 54 & 0 \\
Sanitized command & 0.07 s & 0.72 s & 222 & 0 \\
Sanitized path & 0.05 s & 0.65 s & 129 & 0 \\
Semantic no-op edit & 0.38 s & 0.00 s & 166 & 1 \\
\bottomrule
\end{tabular}
\end{table}

LiveSafeRust consumes the same SAF JSON and checks that diagnostics agree with
SAF's direct verdict. In the semantic no-op run, the LLVM semantic hash is
unchanged after the cache-key normalization above, SAF is not re-invoked, and LiveSafeRust republishes the cached
diagnostic in 0.57 ms. In the sanitizer run, SAF is re-invoked, the finding
disappears, and incremental orchestration recomputes 218 of 222 functions. Five
repeat runs of the six non-cache minimal cases produce SAF median times between
0.53 s and 0.63 s, with IQRs between 0.01 s and 0.06 s.

\section{Comparison}

We ran the minimized \texttt{Command::new} case through existing Rust tooling.
Clippy with \texttt{-W clippy::all -W clippy::pedantic -W clippy::nursery}
emitted no warning. A rust-analyzer LSP probe observed zero diagnostics for an
append-comment edit. CodeQL 2.24.2 created Rust databases and ran the explicit
\texttt{rust-security-extended.qls} suite. The suite contains 27 rules and
produced zero alerts on both our command-injection and path-traversal minimal
cases. GitHub's Rust query reference lists path, SQL, SSRF, crypto, pointer, and
logging queries, but not a built-in command-injection query for
\texttt{std::process::Command}~\cite{codeql-rust}. MIRAI was not installed
locally; its README positions it as a MIR abstract interpreter that can check
security properties, but not as a live LSP diagnostic stream~\cite{mirai}.

\begin{table}[h]
\caption{Existing-tool comparison on the minimized Rust security cases.}
\label{tab:comparison}
\begin{tabular}{lccccc}
\toprule
Tool & Cross-proc. & Live & Security & Incr. & Found \\
\midrule
rust-analyzer & -- & yes & -- & yes & -- \\
Clippy & -- & yes & -- & yes & -- \\
CodeQL Rust & yes & -- & yes & partial & -- \\
MIRAI & yes & -- & partial & -- & n/m \\
LiveSafeRust & yes & yes & yes & yes & yes \\
\bottomrule
\end{tabular}
\end{table}

\section{Detection Sanity}

The sanity suite uses seven 10--30 line Rust programs. We do not claim broad
precision or recall from seven cases; the point is to show that patched SAF and
LiveSafeRust agree on positive and negative cases across two vulnerability
families.

\begin{table}[h]
\caption{Minimal Rust confusion matrix.}
\label{tab:confusion}
\begin{tabular}{lllll}
\toprule
Case & Expected & SAF & LiveSafeRust & Verdict \\
\midrule
\texttt{Command::new} & bug & finding & finding & TP \\
\texttt{Command::arg} & bug & finding & finding & TP \\
\texttt{File::open} & bug & finding & finding & TP \\
RUSTSEC-2024-0350 min. & bug & finding & finding & TP \\
Clean print & clean & none & none & TN \\
Sanitized command & clean & none & none & TN \\
Sanitized path & clean & none & none & TN \\
\bottomrule
\end{tabular}
\end{table}

Aggregate counts are TP=4, FP=0, FN=0, and TN=3. The raw artifacts are saved
under \texttt{experiments/livesaferust/out/saf/r11/} and
\texttt{experiments/livesaferust/out/saf/r12/}.

\section{Related Work}

LiveSafeRust builds directly on the query-based architecture of
rust-analyzer, which uses Salsa for incremental and on-demand computation over
the editor's changing program model~\cite{rustanalyzer,salsa}. Our contribution
is not a new IDE database, but a security-analysis query layer downstream of
the existing Rust frontend.

Compositional analyzers such as Infer show that procedure summaries can make
deep static checks practical in industrial workflows~\cite{infer}. LiveSafeRust
uses the same broad idea of caller-visible summaries, but targets LSP latency
and Rust edit events rather than code-review or CI pipelines.

LLVM-level value-flow frameworks such as SVF and PhASAR provide powerful
interprocedural analyses over compiled IR~\cite{svf,phasar}. LiveSafeRust is
complementary: it is a source/HIR-side frontend intended to surface early
diagnostics before a batch LLVM analysis is available.

Datalog-based systems including CodeQL, Souffle, and Doop demonstrate the value
of declarative program facts and scalable analysis solvers~\cite{codeql,souffle,doop}.
Recent work on incremental CodeQL shows that production security analyses can
benefit from reuse across changes~\cite{inc-codeql}. LiveSafeRust shares that
incremental goal, but focuses on the finer editor loop and on aligning security
facts with rust-analyzer's Salsa invalidation.

\section{Threats To Validity}

The synthetic prototype does not model macros, traits, generics, closures,
async lowering, MIR, or ownership-sensitive unsafe behavior. The real-crate
syn adapter resolves only 44 of 368 call facts in duct; the remaining 324
unresolved calls expose the ceiling of a syn-only frontend and motivate
rust-analyzer HIR integration. LiveSafeRust orchestration timings are
JavaScript microbenchmarks; SAF wall times are reported separately from Docker.
Rust FFI LLVM IR generated with \texttt{rustc 1.79}/LLVM 18 is accepted by SAF
and produces a real finding, and the patched LLVM-22 path now handles minimized
\texttt{std::process::Command} cases. This is still a meaningful subset rather
than full Rust coverage: async lowering, macro-heavy crates, trait-object-heavy
dispatch, richer standard-library protocols, and ownership-sensitive unsafe
behavior remain untested. \texttt{rustc -> LLVM IR} generation is also an
edit-path cost; future work should use incremental IR generation or direct
rust-analyzer HIR to SAF AIR lowering. The method design covers unresolved calls
conservatively, but that policy is not yet implemented in rust-analyzer. The
external baselines cover one real crate and minimized command/path traversal
cases; the rust-analyzer LSP probe produced only empty diagnostics, so it is not
yet a nonempty semantic-diagnostic latency baseline.

\section{Conclusion}

LiveSafeRust proposes a path for moving SAF-derived security findings into the
Rust edit loop. The central observation is that local diagnostics can be
refreshed through body facts and direct callee summaries, while propagation to
ancestors can stop at summary equality. We patched SAF to support a meaningful
subset of \texttt{rustc}-emitted \texttt{std::process::Command} IR and showed an
end-to-end Rust -> SAF -> cached LSP diagnostic path. The remaining task is to
replace the sidecar with rust-analyzer HIR/AIR lowering so the same orchestration
can scale beyond minimized cases.

\begin{thebibliography}{9}

\bibitem{rustanalyzer}
rust-analyzer contributors.
\newblock Architecture.
\newblock \url{https://rust-analyzer.github.io/book/contributing/architecture.html}.

\bibitem{salsa}
Niko Matsakis and Salsa contributors.
\newblock Salsa: incremental recomputation.
\newblock \url{https://rustc-dev-guide.rust-lang.org/queries/salsa.html}.

\bibitem{infer}
Cristiano Calcagno et al.
\newblock Moving Fast with Software Verification.
\newblock In \emph{NASA Formal Methods}, 2015.

\bibitem{svf}
Yulei Sui and Jingling Xue.
\newblock SVF: Interprocedural Static Value-Flow Analysis in LLVM.
\newblock In \emph{Compiler Construction}, 2016.

\bibitem{phasar}
Philipp Dominik Schubert, Ben Hermann, and Eric Bodden.
\newblock PhASAR: An Inter-Procedural Static Analysis Framework for C/C++.
\newblock In \emph{TACAS}, 2019.

\bibitem{codeql}
GitHub.
\newblock About CodeQL.
\newblock \url{https://codeql.github.com/docs/codeql-overview/about-codeql/}.

\bibitem{codeql-rust}
GitHub.
\newblock Rust queries for CodeQL analysis.
\newblock \url{https://docs.github.com/en/code-security/reference/code-scanning/codeql/codeql-queries/rust-built-in-queries}.

\bibitem{mirai}
Endor Labs.
\newblock MIRAI: Rust mid-level IR Abstract Interpreter.
\newblock \url{https://github.com/endorlabs/MIRAI}.

\bibitem{rustsec-gixfs}
RustSec Advisory Database.
\newblock RUSTSEC-2024-0350: gix-fs traversal outside working tree enables
arbitrary code execution.
\newblock \url{https://rustsec.org/advisories/RUSTSEC-2024-0350.html}.

\bibitem{inc-codeql}
Tamas Szabo.
\newblock Incrementalizing Production CodeQL Analyses.
\newblock In \emph{ESEC/FSE}, 2023.

\bibitem{souffle}
Herbert Jordan, Bernhard Scholz, and Pavle Subotic.
\newblock Souffle: On Synthesis of Program Analyzers.
\newblock In \emph{CAV}, 2016.

\bibitem{doop}
Martin Bravenboer and Yannis Smaragdakis.
\newblock Strictly Declarative Specification of Sophisticated Points-to Analyses.
\newblock In \emph{OOPSLA}, 2009.

\end{thebibliography}

\end{document}
